Analyse gender in music psychology from journal publication data

### 2025-09-19 manuscript/manuscript.qmd
Added a table and rewrote conclusions, polished keywords, and highlighted issues that need attention

keywords          : "gender, transparency, music psychology, open science, meta-research"
wordcount         : "5555"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
csl: apa7.csl
class             : "man"
bibliography: references.bib
format:
  html: default
---

@@

Research across several scientific disciplines show that women tend to be less represented as authors compared to men [@banks2025women; @son2022scientific]. As the topics, approaches, and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%), but not equal. In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067). Encouragingly, however, studies looking at various year ranges between 1960 until 2024 show a general trend that the ratio of women to men in authorships is improving [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. Yet, there is evidence that this improvement is plateauing [@jemielniak2025gender]. Further, although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@bruck2023bibliometric; @gonzalez2020women; @rock2021quantifying; @shah2021gender]. There could be several reasons for such a disparity: two key aspects that are typically explored in such <article analysis> papers: geographical locations and keyword/topics.

@@

Gender attribution was initially based on first names. We recognise that treating gender as a binary category is inherently problematic. Gender is a complex and multidimensional social construct. However, consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006] and that non-binary authorship generally is less than 0.1% [@son2022scientific], we adopt a binary classification—male and female—for analytical purposes and assume that first names allow for a reasonable, though imperfect, attribution of gender. We used the _genderize API_ [@wais2006], which predicts gender from first names and can be supplemented with country information derived from author affiliations to improve accuracy. This method resolved the gender of 89.3% of authors with a probability greater than 0.90. Only `89` names had a low attribution probability (< 0.55). Unattributed cases were then checked manually, resulting in `185` manual corrections. After this process, `32` names remained ambiguous and `27` were unknown—some likely due to data entry errors in Scopus (e.g. only initials or surnames). These `59` cases were excluded from the dataset. It is likely that the due to challenge of attributing gender accurately not all the gender attributions are correct. Although genderize.io has been shown to achieve 96.6% accuracy in a diverse multinational test database without using a country of origin information -- 98% with the information included -- it also underperforms (82% accuracy) in Asian names [@vanhelene2024inferring]. However, the error rates of these processes have been previously been shown to be non-biased, i.e. showing similar number of mistakes for both genders [@sebo2021performance;@vanhelene2024inferring].

@@

The overall OR for women as first authors was 1.408 [95% CI: 1.284–1.544], indicating that the odds of first authorship are 41% higher for females compared to the male authors (Fisher's exact test _p_ < .001). For coauthorship positions, the OR was 0.997 [0.913–1.090], suggesting no substantial gender difference (_p_ = .96). In contrast, the odds ratio for women as last authors was 0.733 [0.667–0.806], indicating that the odds of last authorship are 27% lower for women compared to men and significantly different (_p_ < .001). Being a single author has odds ratio of 0.905 [0.778-1.052] for females, which suggests that while women are represented with relatively more single authored papers than men, the difference is not significant (_p_ = .21).

@@

To explore the potential differences within music psychology topics of choice between female and males, we analysed the keywords in the articles (15,316 in total, 6,730 unique). To reduce variant keywords, we manually simplified the variants (e.g., "arts in health", "healthy music use", "health musicking" were converted to "music health" and "music cognition" to "cognition" and "music perception" to "perception") and eliminated single use of the term "music". We aggregated the counts of keywords across all authors in the articles and Figure 4 shows the odds ratios for female authors to be associated (in any author position) with the 40 most frequently used keywords. The keywords signifying related to children, development, emotion regulation, well-being, music performance anxiety, motivation, expertise, music listening, mental health, and music therapy show significantly higher odds ratios (OR $\ge$ 1.50) for females in comparison to overall gender ratios in the data. At the opposite end of this continuum, topics such as music analysis, harmony, machine learning, computation, cognition, and performance expression show the opposite trend, suggesting that the articles with these keywords tend to have overall more prevalent male authorships (female OR $\le$ 0.75). There is much nuance and detail in the data that would warrant separate analyses; however, the purpose here is simply to provide reference points for comparing gender distributions of authorship with other disciplines, where thematic analyses have already been conducted (e.g., in sub-disciplines of psychology).

@@

In music psychology, the distribution of authors by gender remains unequal, with women representing 40.2% of authors. This proportion is higher than in musicology (34.8% [@eerola2026inprep]) and higher than in psychology more broadly, where women account for 33.2% of authors [@huang2020]. In terms of authorship roles, notable differences emerge. Women are relatively more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. This supports previous research showing that there is a positive trend with female representation increasing [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. In contrast, women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. This corresponds to the situation in other academic fields such as medicine [@bruck2023bibliometric; @shah2021gender], psychology [@gonzalez2020women], and biology [@rock2021quantifying]. Given that last authorship in the social sciences is not universally defined but is increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution], this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science]. This suggests that the more 'leadership' author position is more likely to go to the male author.

@@

->  (potential reasons: women faculty perform more university/campus/departmental services @guarino2017faculty )

Turning to potential drivers of these gender differences, geographical and cultural factors appear to play a major role. Our summaries of gender patterns across continents and countries indicate clear cross-national variation. Large Commonwealth nations, namely Australia, Canada, and the United Kingdom, seem to provide a cultural climate more supportive of women in research, as reflected in the relatively high odds ratios of women appearing as authors (in any authorship position) in music psychology. This pattern, however, is not universal across all countries examined [@chan2020]. Prior work shows that women’s research success is positively correlated with national indicators of gender equity [@chan2020]. This is likely to be related to a career length, which strongly correlates with overall productivity. Women exhibit higher dropout rates than men, particularly during the 5–10 years following their first publication [@huang2020], but policies geared towards equity and support for career breaks and caring responsibilities may well contribute towards the observed country-specific differences. Nevertheless, prior analyses demonstrate that even when women constitute a smaller share of researchers within a discipline or country, their contributions are often more impactful than those of their male colleagues [@chan2020]. When examining countries where women have significantly lower odds of authorship compared to men (Japan, Belgium, Austria, Sweden, and Turkey), it is not straightforward to conclude that these nations fall consistently at the lower end of gender equality rankings. The notable exception is Turkey, which is positioned lower according to the Gender Inequality Index [@UNDPHDR2024].

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on (health)care and nurturing [@bruck2023bibliometric; @gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women]. In a similar trend in computational linguistics, woman are more likely to publish on speech, social and conversational topics than males, who are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he]. It perhaps might be interesting to note that 'emotion' has an equal balance between men and women. Although this may come as a surprise hen coming from different disciplines, music emotion can be a relatively 'analytical' aspect in music psychology, where emotion from music is attempted to be decoded in a more analytical way (toiviainen papers?), as well as being used in a more 'emotional' sense. this is the same for the keyword 'expression'. Although on first glance, it might fit the stereotype that this keyword might be more related to females, the 'expressivity' is again related to a more analytical coding aspect <cite Wanderley/thompson etc.>

In summary, music psychology has similar groupings of topic for female and male authors as in other disciplines of psychology and medicines. Nonetheless, it should a positive trend. Below we also discuss how this information can help benefit the general field towards more equity with academia.

<!-- I did not know what these references signaled, so temporarily disabled them: -->
<!-- [@ashmos2011gendering; @sebo2020gender; @thelwall2019gender], -->

@@

There was small but significant effect of gender on citations that favoured men when the citations were aggregated all authors in the papers. There was no, however, differences in citations for female first or last authors. The aggregated results are in line with previous findings in psychology that favored men [@gonzalez2020women]. While the present analysis cannot determine the underlying causes of the observed trends, we can point to potential factors that may help explain why this discipline appears to be more equitably distributed by gender for the first authors than, for example, psychology. Female role models (past and present) exist in music psychology; many of the founding members of the societies have been women (e.g., Diana Deutsch for ICMPC in 1989, Irène Deliège for ESCOM in 1991, and similarly many founding editors of the journals have been women (e.g., Diana Deutsch for _Music Perception_ 1983-1995, and Irène Deliège for Musicae Scientiae 1997-2009). However, this does not fully explain the rise of female first authors over the past 15 years, nor is it consistent with the reverse pattern observed in the last authorship position. A cautious interpretation is that the increase in first authorship may signal the discipline becoming more gender-balanced, but this shift has not yet extended to last authorship positions, which still reflect either career longevity or barriers to women attaining senior academic positions.

While the analysis of gender of the authorships across the discipline-specific journal is relatively simple summary in the present study, there are more elaborate ways into obtaining insights into the gendered career progression in this disciple; for instance, future work could analyse the precarity of career options as judged from the years of authorship attributions [@lundine2018]. One could also analyse the last author positions in more detail using direct contacts to the authors (interviews and surveys) and try to pin down the factors contributing to the barriers experienced by women to take up senior author positions.

Current findings present a more optimistic outlook of female authors in comparison what the prevalence of female academics or professors in academia are [@bhopal2021competing;@harris2025gender;@Filandri03082021]. While the results are comparable to psychology in general, musicology (and the closely associated disciplines such as music theory, popular music and ethnomusicology) tend to have less equal division of gender in overall authorship; a recent analysis of the last 25 years of articles in 15 most eminent musicology journals amounting to 3,515 articles suggests that female authors are far more infrequent (34.8% [33.3%-36.3%]) than males (65.2% [63.7%-66.7%]) [@eerola2026inprep]. The present findings should be used in conjunction of ...

>>>>

keywords          : "gender, transparency, music psychology, open science, meta-research"
wordcount         : "5578"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
csl: apa7.csl
class             : "man"
bibliography: references.bib
format:
  html: default
---

@@

Research across several scientific disciplines show that women tend to be less represented as authors compared to men [@banks2025women; @son2022scientific]. As the topics, approaches, and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%), but not equal. In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067). Encouragingly, however, studies looking at various year ranges between 1960 until 2024 show a general trend that the ratio of women to men in authorships is improving [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. Yet, there is evidence that this improvement is plateauing [@jemielniak2025gender]. Further, although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@bruck2023bibliometric; @gonzalez2020women; @rock2021quantifying; @shah2021gender]. There could be several reasons for such a disparity: two key aspects that are typically explored in such **article analysis [TE: did you think citing specific studies here?]** papers: geographical locations and keyword/topics.

@@

Gender attribution was initially based on first names. We recognise that treating gender as a binary category is inherently problematic. Gender is a complex and multidimensional social construct. However, consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006] and that non-binary authorship generally is less than 0.1% [@son2022scientific], we adopt a binary classification—male and female—for analytical purposes and assume that first names allow for a reasonable, though imperfect, attribution of gender. We used the _genderize API_ [@wais2006], which predicts gender from first names and can be supplemented with country information derived from author affiliations to improve accuracy. This method resolved the gender of 89.3% of authors with a probability greater than 0.90. Only `89` names had a low attribution probability (< 0.55). Unattributed cases were then checked manually, resulting in `185` manual corrections. After this process, `32` names remained ambiguous and `27` were unknown—some likely due to data entry errors in Scopus (e.g. only initials or surnames). These `59` cases were excluded from the dataset. It is likely that the due to challenge of attributing gender accurately not all the gender attributions are correct. Although genderize.io has been shown to achieve 96.6% accuracy in a diverse multinational test database without using a country of origin information -- 98% with the information included -- it also underperforms (82% accuracy) in Asian names [@vanhelene2024inferring]. However, the error rates of these processes have been previously been shown to be non-biased, i.e. showing similar number of mistakes for both genders [@sebo2021performance;@vanhelene2024inferring;@thelwall2019gender].

@@

A summary of the authorship types and positions are given in Table 1. While the single-authored papers are in minority (8.6%) in this sample overall, there does not seem to be a significant difference between men and women authoring such articles (_p_ = .21); 38.05% of single-authored papers are women with an odds ratio of 0.90 [95% 0.78-1.05] suggesting no clear gender difference despite the figures leaning towards higher frequency of male authors. The overall OR for women as first authors was 1.41 [95% CI: 1.28–1.54], indicating that the odds of first authorship are 41% higher for females compared to the male authors (Fisher's exact test _p_ < .001). For coauthorship positions, the OR was 1.00 [0.91–1.09], suggesting no substantial gender difference (_p_ = .96). In contrast, the odds ratio for women as last authors was 0.73 [0.67–0.81], indicating that the odds of last authorship are 27% lower for women compared to men and significantly different (_p_ < .001).

```{r}
#| id: table1
#| eval: true
#| echo: false
#| warning: false
#| message: false
#| output: asis
d <- read.csv('../data/authorship_counts_table.csv',header=TRUE)
colnames(d)[1]<-"Author Type"
colnames(d)[4]<-"Female %"
colnames(d)[6]<-"OR (CI lower)"
colnames(d)[7]<-"OR (CI higher)"
print(knitr::kable(d,digits=2,caption="Summary of authorship types across gender."))
```

@@

To explore the potential differences within music psychology topics of choice between female and males, we analysed the keywords in the articles (41,438 in total, 5,960 unique). To reduce variant keywords, we manually simplified the variants (e.g., "arts in health", "healthy music use", "health musicking" were converted to "music health" and "music cognition" to "cognition" and "music perception" to "perception") and eliminated single use of the term "music". We aggregated the counts of keywords across all authors in the articles and Figure 4 shows the odds ratios for female authors to be associated (in any author position) with the 40 most frequently used keywords. The keywords signifying related to children, development, emotion regulation, well-being, music performance anxiety, motivation, expertise, music listening, mental health, and music therapy show significantly higher odds ratios (OR $\ge$ 1.50) for females in comparison to overall gender ratios in the data. At the opposite end of this continuum, topics such as music analysis, harmony, machine learning, computation, cognition, and performance expression show the opposite trend, suggesting that the articles with these keywords tend to have overall more prevalent male authorships (female OR $\le$ 0.75). There is much nuance and detail in the data that would warrant separate analyses; however, the purpose here is simply to provide reference points for comparing gender distributions of authorship with other disciplines, where thematic analyses have already been conducted (e.g., in sub-disciplines of psychology).

@@

In music psychology, the distribution of authors by gender remains unequal, with women representing 40.2% of authors. This proportion is higher than in musicology (34.8% @eerola2026inprep) and higher than in psychology more broadly, where women account for 33.2% of authors [@huang2020]. In terms of authorship roles, notable differences emerge. Women are relatively more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. This supports previous research showing that there is a positive trend with female representation increasing [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. In contrast, women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. This corresponds to the situation in other academic fields such as medicine [@bruck2023bibliometric; @shah2021gender], psychology [@gonzalez2020women], and biology [@rock2021quantifying]. Given that last authorship in the social sciences is not universally defined but is increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution], this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science]. This suggests that the more 'leadership' author position is more likely to go to the male author.

@@

**->  (potential reasons: women faculty perform more university/campus/departmental services @guarino2017faculty )
[TE: is this supporting the idea that women are publishing more access than men? Or is this more about the lack of senior positions, which are probably not going to be achieved with internal service?]**

Turning to potential drivers of these gender differences, geographical and cultural factors appear to play a major role. Our summaries of gender patterns across continents and countries indicate clear cross-national variation. Large Commonwealth nations, namely Australia, Canada, and the United Kingdom, seem to provide a cultural climate more supportive of women in research, as reflected in the relatively high odds ratios of women appearing as authors (in any authorship position) in music psychology. This pattern, however, is not universal across all countries examined [@chan2020]. Prior work shows that women’s research success is positively correlated with national indicators of gender equity [@chan2020]. This is likely to be related to a career length, which strongly correlates with overall productivity. Women exhibit higher dropout rates than men, particularly during the 5–10 years following their first publication [@huang2020], but policies geared towards equity and support for career breaks and caring responsibilities may well contribute towards the observed country-specific differences. Nevertheless, prior analyses demonstrate that even when women constitute a smaller share of researchers within a discipline or country, their contributions are often more impactful than those of their male colleagues [@chan2020]. When examining countries where women have significantly lower odds of authorship compared to men (Japan, Belgium, Austria, Sweden, and Turkey), it is not straightforward to conclude that the lower proportion of female author is related gender equality rankings. The notable exception is Turkey, which is positioned lower according to the Gender Inequality Index [@UNDPHDR2024].

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on (health)care and nurturing [@bruck2023bibliometric; @gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women]. In a similar trend in computational linguistics, woman are more likely to publish on speech, social and conversational topics than males, who are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he]. It perhaps might be interesting to note that 'emotion' has an equal balance between men and women. **Although this may come as a surprise coming from different disciplines, music emotion can be a relatively 'analytical' aspect in music psychology, where emotion from music is attempted to be decoded in a more analytical way (top 3 authors for the keyword "emotion" are majority male, Saarikallio, Juslin, and Schubert, but "emotion regulation" top three is fully female), as well as being used in a more 'emotional' sense. This is the same for the keyword "expression". Although on first glance, it might fit the stereotype that this keyword might be more related to females, the "performance expression"  is again related to a more analytical coding aspect (Marc Leman, Laura Bishop, and Roger Chaffin as top 3 authors).**

In summary, music psychology has similar groupings of topic for female and male authors as in other disciplines of psychology and medicines. Below we also discuss how this information can help benefit the general field towards more equity with academia.

<!-- I did not know what these references signaled, so temporarily disabled them: -->

<!-- @ashmos2011gendering : women publish more qualitative studies than men -->

<!-- @sebo2020gender : women were more likely to publish qualitative studies and to publish in primary healthcare journals -->

<!-- @thelwall2019gender : women are likely to use exploratory and qualitative methods rather than quantitative methods. -->

@@

There was small but significant effect of gender on citations that favoured men when the citations were aggregated all authors in the papers. There was no, however, differences in citations for female first or last authors. The aggregated results are in line with previous findings in psychology that favored men [@gonzalez2020women]. While the present analysis cannot determine the underlying causes of the observed trends, we can point to potential factors that may help explain why this discipline appears to be more equitably distributed by gender for the first authors than, for example, psychology. Female role models (past and present) exist in music psychology; many of the founding members of the societies have been women (e.g., Diana Deutsch for ICMPC in 1989, Irène Deliège for ESCOM in 1991, and similarly many founding editors of the journals have been women (e.g., Diana Deutsch for _Music Perception_ 1983-1995, and Irène Deliège for _Musicae Scientiae_ 1997-2009). However, this does not fully explain the rise of female first authors over the past 15 years, nor is it consistent with the reverse pattern observed in the last authorship position. A cautious interpretation is that the increase in first authorship may signal the discipline becoming more gender-balanced, but this shift has not yet extended to last authorship positions, which still reflect either career longevity or barriers to women attaining senior academic positions.

While we present analysis of gender of the authorships across the discipline-specific journal in a relatively simple manner here, there would be more elaborate ways into obtaining insights into the gendered career progression in music psychology; for instance, future work with the same dataa could analyse the precarity of career options as judged from the years of authorship attributions [see @lundine2018]. One could also analyse the last author positions in more detail using direct contacts to the authors (interviews and surveys) and try to identify the factors contributing to the barriers experienced by women to take up senior author positions.

Current findings from authorship counts in music psychology present a more optimistic outlook of female authors in comparison what the prevalence of female academics or professors in academia in general are [@bhopal2021competing;@harris2025gender;@Filandri03082021]. While the results are broadly comparable to psychology as a whole, musicology and its closely related disciplines such as music theory, popular music studies, and ethnomusicology tend to show less balanced gender representation in overall authorship. Gender, however, is only one dimension of equity; racial, ethnic, geographical, institutional, and other inequalities also merit attention in music psychology. Although women remain underrepresented in terms of absolute authorship counts in specialist music psychology journals, even a simple tally of contributions can help open discussion and provide evidence of progress toward greater inclusion and equality. To interpret these patterns more fully, one must examine how music psychology operates through its journals, conferences, training programmes, and curricula, as well as how the discipline presents itself both in countries with a long-standing tradition in the field and in regions where it has yet to become firmly established.
